Think deep for reef resilience

### 2025-08-17 paper/paper_draft.qmd
Various proof-reading changes and improvements

Long-term reef viability requires not only coral survival and recolonization, but also
maintenance of the physical reef structure through positive carbonate budgets. Scleractinian
corals build the primary habitat framework, but reef persistence depends on the balance
between carbonate accumulation [@perry2015] and loss through bioerosion, physical damage
[@kenyon2023], and dissolution [@eyre2014; @eyre2018]. This balance is influenced by coral
cover and community complexity [@perry2012; @desbiens2021], environmental processes, and
ocean chemistry [@perry2015; @eyre2018]. For Great Barrier Reef mid and outer shelf reefs,
maintaining positive carbonate budgets requires approximately 17% live coral cover
[@desbiens2021].

@@

![Map of the Great Barrier Reef region, including GBRMPA Management areas (polygons) and all
considered reefs (points). Colour-coded areas outline GBR marine park management regions,
where orange represents the Far Northern management region, green represents
Cairns/Cooktown, red covers Townsville/Whitsunday and blue covers Mackay/Capricorn
[@GBRMPA2007]. Reef centroids are marked individually for visual clarity, and are coloured
according to its bioregion [@GBRMPA2024]. All 2429 reefs included in this investigation are
represented in this figure. Annotations represent major coastal settlements along the
GBR.](../figs/region_map.png){#fig-region-map fig-align="center" width="14.8cm"}

@@

These GCMs were selected based on evaluation conducted by Grose et al., [-@grose2023],
indicating their ability to recreate historical atmospheric temperature and precipitation
conditions in the Australasian region. They span a range of Equilibrium Climate Sensitivity
(ECS) values within the Intergovernmental Panel on Climate Change (IPCC) assessed likely
range. The IPCC defines ECS as the long term warming caused by a doubling of carbon dioxide
above its pre-industrial concentration [@forster2021]. While this assessment does not
consider oceanic data, Grose et al. [-@grose2023] provide a region-specific selection
mechanism which we use as an indicator of general performance of GCMs for the region. An
overview of the selected GCMs are provided in @tbl-gcms. Note here that Grose et al.,
[-@grose2023] assessed GFDL-ESM4, whereas Dixon et al., [-@dixon2022; -@dixon2023] provides
data based on GFDL-CM4. This substitution is assumed appropriate given they share the same
fundamental representation of the physical climate system.

Projections using Shared Socio-economic Pathway (SSP) 2-4.5 was used to represent the
plausible emission pathway. It represents a moderate emissions scenario with continued CO2
emissions comparable to present day until 2050, after which emissions gradually decline
[@hausfather2020; IPCC -@climate2023]. The selected GCMs represent 2.5°C to 4.66°C of global
atmospheric warming in the 2081-2100 period relative to 1850-1900. The use of these GCMs
therefore allow for a range of possible climate futures to be explored, capturing the
uncertainty in future responses to CO2 increases (@sfig-ecs; @fig-methods-dhws; [@grose2023;
@shepherd2018]) and enables the assessment of model projected coral reef behaviour over a
broad range of independently derived climate futures [@abramowitz2019].

Monthly mean SST climatological baselines, for each 1km^2^ pixel, were calculated from the
daily SST data following NOAA methodology [@liu2017; @skirving2020]. For each pixel of SST
data, the Monthly Mean SST climatology was calculated by computing the mean SST value for
each calendar month (January through December) across the 28-year baseline period from 1985
to 2012. For each month separately, this climatological baseline was then adjusted to center
at year 1988.2857, which attempts to account for any long-term trends within the baseline
years themselves (see Skirving et al. [@skirving2020]). The adjustment ensures that the
climatology represents conditions at a specific temporal reference point rather than being
influenced by gradual warming or cooling trends that may have occurred during the 1985-2012
period. The resulting monthly climatologies serve as a standardized reference against which
current SST conditions can be compared to identify thermal anomalies and assess coral
bleaching risk, following the methodology established by Skirving et al. [-@skirving2020].

@@

::: {#tbl-gcms}
| Global Climate Model source | ECS value (°C)     | Interpreted relative severity | Results found in |
|-----------------------|-----------------------|------------------------|-----------------------|
| ACCESS-CM2                  | 4.66 [@grose2023]  | High                          | [@grier2025]     |
| EC-Earth3-Veg               | 4.30 [@wyser2020]  | High                          | [@grier2025]     |
| GFDL-CM4                    | 4.1 [@sentman2025] | High                          | [@grier2025]     |
| ACCESS-ESM1-5               | 3.87 [@ziehn2020]  | Intermediate                  | Main text        |
| NorESM2-MM                  | 2.5 [@seland2020]  | Low                           | [@grier2025]     |

: Information for each GCM used in this investigation including Equilibrium Climate
Sensitivity (ECS) value, interpreted relative severity and where results for an individual
GCM results are found. GCMs in context with assessed plausible ranges of ECS values are
displayed in [@sfig-ecs]. ACCESS-ESM1-5 has an ECS closest to the midpoint of the
very-likely ECS range and was taken as an intermediate model with results displayed in the
main text.
:::

@@

An exploratory scenario modelling approach is adopted for this investigation. The method
involves exploring a wide range of plausible futures under different assumptions of
environmental (climatic) conditions and ecological factors. Exploratory scenario modelling
embraces uncertainty as an inherent characteristic of complex environmental systems
[@lempert2010] and is a core methodology underpinning the development of robust and adaptive
management strategies [@haasnoot2013].

Exploratory modelling approaches have gained favour in recent years as precise long-term
predictions of future conditions are often impossible in complex systems like coral reefs.
Multiple interacting and compounding sources of uncertainty, and a lack of broad agreement
as to their likelihood, create inherently uncertain outcomes. Rather than attempting to
forecast specific outcomes, exploratory modelling aims to systematically explore many
possible futures to identify patterns and, in management contexts, robust strategies that
perform well across those diverse conditions, as well as conditions in which such strategies
may fail. Such an exploratory approach lends itself well to the purpose of the
investigation. Specifically, to assess the role of DHW, connectivity, and depth to identify
the environmental and reef characteristics that support structural resilience.

@@

To apply exploratory scenario modelling, the approach employed a modelling framework
involving two separate systems: ADRIA (the Adaptive Dynamic Reef Interventin Algorithms
platform) and CoralBlox. ADRIA is a multi-criteria decision support platform designed to
help reef managers address climate change challenges in coral reef systems [@iwanaga2023].
The ADRIA platform provides functionality for scenario generation and analysis as well as
providing the technical infrastructure to couple environmental and ecosystem
representations. CoralBlox is a computationally efficient coral population model. It
simulates functional/morphological groups of corals over its life cycle, competing for
available space over time [@ribeirodealmeida2024]. The coupled framework is referred to as
ADRIA-CoralBlox.

The coral demographic representation and associated model implementation are further
detailed in the supplementary material. Of important note, however, is that the
representation of bleaching mortalities follow Bozec et al., [-@bozec2022] albeit with a
scaling coefficient to incorporate the amelioration effect that depth may have, following
Baird et al., [@baird2018]. The scaling reduces the proportion of coral colonies impacted by
high heat stress as depth increases. Tropical cyclones and impact of wave activity on reefs
were not included in this investigation.

@@

\*\* need information on GBR1 connectivity used \*\*

Connectivity data from 24 days in total were used in this investigation to estimate coral
larval flows. These data span six years from 2015 to 2020. The years 2015, 2016, 2018 and
2019 are each represented by connectivity matrices covering spawning periods over three days
in each year (12 days in total), while 2017 and 2020 are each represented by six spawning
days each, coming to a total of 24 days of connectivity matrices. Indicative connectivity
strengths are implemented in ADRIA-CoralBlox as the mean of all available data for each
location (i.e., mean of all 24 spawning days). While the connectivity strengths do not vary
from year to year within each simulation, they do vary between reefs which are of varying
sizes, providing sufficient heterogeneity to capture spatial patterns in larval dispersal
and settlement dynamics. The spatial variability also enables investigations into the
influence of connectivity strength and its role in driving recovery processes.

@@

-   $S$ is the sum of outgoing and incoming connections, indicating the total connectivity
    strength.

@@

The time series clustering was applied over the time frame between 2025 and 2075. Spatially,
the clustering analyses were conducted for each GBR management area and at the scale of each
reefal bioregion (defined in GBRMPA [-@GBRMPA2024]). Analyses were also performed at the
whole-of-GBR scale but were found to lead to identical conclusions about characteristics
influencing cluster cover compared to management area and bioregion scale analyses, and so
are not shown here. To compare time series clustering results across GCMs, management area
scale clustering results are used to preserve coarse scale latitudinal differences between
regions, while maintaining ease of comparison and visualisation. These comparisons include
quantification of differences between low and high cover clusters across management regions
under each GCM using the CID between cluster median time series. To assess how different
reef characteristics influence time series clustering at a finer spatial scale, bioregion
results are used to capture patterns occurring within individual groups of reefs. Results of
all analyses at all clustering scales can be found at Grier and Iwanaga [-@grier2025] for
those interested.

@@

Partial Dependence Analysis was conducted to understand the relationship between individual
features and time series cluster predictions using a trained Random Forest (RF) classifier
[@friedman2001]. The analysis utilised the reef characteristics of interest as the
predictors, including the median reef depth, mean DHW for the given GCM, total connectivity
strength, and the outgoing-to-incoming connectivity ratio. The full dataset consisted of the
time series cluster assignment of each of the 2429 reefs under the five GCMs explored, for a
total of 12,145 samples. The dataset was randomly split into training (60%) and testing
(40%) subsets to ensure model generalisability. The RF classifier achieved a 99% accuracy on
the training set and 66% on the test set.

While the 99% accuracy suggests the classifier is overfit to the training data, the 66% test
accuracy indicates a meaningful relationship between modelled reef characteristics and their
assigned clusters has been captured. A 3-fold cross-validation was also conducted, arriving
at a similar 67% accuracy rate. The consistent level of accuracy across test and
cross-validation demonstrates the generalised performance of the RF classifier is reliable.
Furthermore, F1 performance scores for Low, Medium, and High were determined to be 0.7, 0.57
and 0.68 respectively, indicating balanced performance across the cluster categories. Lower
performance with the Medium cluster reflect the inherent difficulty of classifying
transitional states in natural systems. For management applications, this level of
performance provides valuable decision support when combined with expert knowledge and the
model's probabilistic outputs.

Partial dependence analysis was performed by applying the trained model to the test dataset.
Values for individual predictors were systematically (one-at-a-time) varied across a grid of
50 equally spaced points spanning its range, while all other features were left at their
modelled values. The procedure creates (1) a modified test dataset, where the target
predictor was set to the grid value for all samples, (2) generates probabilistic predictions
for all three cluster classes using the trained RF classifier, and (3) the mean predicted
probability for each class was calculated across all test observations. The median was also
assessed (not shown) but was found not to substantially influence the interpretation of the
analysis results. The process isolates the marginal effect of each predictor by averaging
over the natural distribution of all other representative reef characteristics in the test
dataset.

@@

Using coral cover time series for each reef we calculated the number of years each reef
maintains coral cover above a given percent threshold. The threshold represents the level of
live coral cover necessary to maintain a positive carbonate budget (denoted here as
$\theta_{0}$ to indicate the budget equilibrium). Greater number of years above $\theta_{0}$
therefore suggests greater resilience and opportunity for recovery.

@@

The relationship between the number of years above $\theta_{0}$ and median reef depth was
then assessed with Spearman's rank correlation. Spearman's correlation coefficient was
adopted as reef depth and connectivity strength distributions are expected to be non-normal
and highly skewed. Furthermore, reefs are naturally of different types and experience
different conditions, and so the monotonic trends (whether the relationship consistently
increase or decrease) rather than the exact quantitative relationship was of interest for
this analysis.

@@

There were substantial differences in the maintenance of coral cover post 2030 between the
GCMs. Figure [-@fig-gcm-cover-trajectories] illustrates the temporal behaviour of reefs by
their categorisation (Low, Medium and High). Results are displayed at the Management area
scale to illustrate the spatial differences in cover trajectories. Of the GCMs explored,
NorESM2-MM has the lowest Equilibrium Climate Sensitivity (2.5°C) and reefs exhibit
continuing recovery and decline dynamics up to 2075 under this hypothetical climate future.

Unsurprisingly, recovery phases are predominantly seen in the reefs in the High category,
with Medium and Low exhibiting lesser probability of such periods of recovery. Generally,
recovery and increases in coral cover are visible in simulated results up to 2045. Under
NorESM2-MM, the Far Northern Management area were found to have the most similar temporal
behaviour between Low and High categories compared to other Management areas
[@tbl-cluster-cid] due to the low degree of coral cover [@fig-gcm-cover-trajectories].

Intermediate levels of coral cover maintenance can be seen in ACCESS-ESM1-5 (ECS value of
3.87°C) when compared to the other GCMs (of higher and lower ECS values). Under these
intermediate conditions (in relative terms), there is potential for reefs in the High
category to maintain coral cover levels around 20% to 2050. The highest peak DHW levels
occur under GFDL-CM4 between 2045 and 2050, while greater increases in overall DHWs time
series occur under ACCESS-CM2 [@fig-methods-dhws]. Under conditions represented by high ECS
value GCMs - specifically, ACCESS-CM2 (4.66°C), EC-Earth3-Veg (4.30°C) and GFDL-CM4
(4.1°C) - all reefs are largely in decline across all management areas from 2055
[@tbl-cluster-cid].

@@

Comparing across GCMs, it is common for reefs to switch between categories, including from
Low to High and vice versa (@sfig-gcm-bioregion-cluster-assignment). Of 2429 reefs, 20%
(503) of reefs have consistent category assignment across all GCMs. Of the 503 consistently
assigned reefs, 49% are within the Far Northern Management area, with 7%, 13% and 31%
occurring in the Cairns-Cooktown, Townsville-Whitsunday and Mackay-Capricorn Management
areas respectively [@sfig-gcm-cluster-map]. Of the 503 consistently assigned reefs, 32%, 7%
and 61% are clustered into Low, Medium and High clusters respectively. High assignment
variability across GCMs is expected as spatial differences in a given GCM's DHW trajectory
result in reefs being impacted at different relative magnitudes within each Management
areas.

```{julia}
#| echo: false
#| warning: false
#| message: false
#| results: asis
#| label: tbl-cluster-cid
#| tbl-cap: Complexity invariant distance (CID) between low and high cluster median time series in each management area under each GCM. CID values are calculated for percentage cover time series (@fig-gcm-cover-trajectories) and Degree Heating Week (DHW) time series (@sfig-gcm-dhw-trajectories). Complexity invariant distance represents the euclidean distance between time series, corrected by the complexity of the two time series. CID values are used to quantify overall differences between these two cluster levels at management area and GCM levels. CID values for percentage cover and DHW time series are not directly comparable as the time series use different units.

@@

In contrast to the broad temporal behaviour at the Management area scale, the relationship
between reef characteristics and reefs' assigned categories were found to be similar at the
bioregion scale across GCMs. Results are only shown and discussed for ACCESS-ESM1-5, an
intermediate Equilibrium Climate Sensitivity scenario [@sfig-ecs]. Those interested in
results for the other assessed GCMs are referred to Grier and Iwanaga [-@grier2025]. Across
bioregions, reefs in the High category not only had greater levels of coral recovery but
displayed persistence to 2050 [@sfig-access-esm1-5-bioregion-timeseries]. Reefs in the High
category were consistently found to have the greatest median reef depths across bioregions
[@fig-access-esm1-5-bioregion-depth].

@@

In many bioregions reefs in the Low category have the lowest correlation between coral cover
and DHW whereas reefs in the High category were found to have the strongest correlation
(@sfig-access-esm1-5-bioregion-dhwcor). Conversely, there are no consistent patterns across
bioregions when comparing their temporal behaviour with respect to their reef area and mean
of DHWs (@sfig-access-esm1-5-bioregion-dhwmean). There were no consistent differences across
bioregions and GCMs between clusters in their reefs' initial proportion of habitable area
occupied by corals (@sfig-access-esm1-5-bioregion-initialproportion). Across most bioregions
there were not consistent differences between clusters in connectivity metrics such as
source-to-sink connectivity ratio (@sfig-access-esm1-5-bioregion-logsosi) and total
connectivity strength (@sfig-access-esm1-5-bioregion-totalconnectivity).

**\[TI: Note, I'm not sure if this section is interesting enough to keep now... let's
discuss.\]**

@@

A Partial Dependence Analysis was conducted using a Random Forest classifier to identify
reef characteristics that influence its temporal behaviour, and thus its cluster assignment
into Low, Medium, or High. These include, for each of the 2429 reefs assessed, the
indicative median reef depth, the mean DHW experienced, its coral carrying capacity, total
connectivity strength and source-to-sink ratio. Reefs' bioregion membership was excluded
from this analysis due to its categorical nature providing limited interpretable
information, and is further justified by its low feature importance score of 10%, the lowest
of all predictors [-@fig-ts-cluster-pdp, Panel A]. Reef median depth and mean DHW level have
the highest feature importance scores, accounting for 25% and 23% predictive importance
respectively. The carrying capacity of a reef accounts for 14.3% feature importance.
Connectivity strength and source-to-sink ratio both had feature importance scores of \~13%.

Reefs have the greatest probability of being clustered into Low if shallow, and High at
greater depths, with medium cluster reefs exhibiting a lower, and stable, variation in
assignment probability across depth ranges [@fig-ts-cluster-pdp, Panel B]. A large increase
in the probability of a reef being categorised as High can be seen between 10 and 15 meters
depth. This is at the edge of the depth range used in current deployment efforts
[@howlett2021; @randall2021; @whitman2024], but indicates a depth range that could be
adopted to maximise longevity and benefits of, for example, coral seeding activities. With
regard to DHWs, assignment probability is roughly equal across the three categories up to 6
DHW. A sharp delineation can be seen \> 6 DHW when comparing High with other categories,
with lower assignment probability (\~25%) compared to Medium and Low (\~40%)
[@fig-ts-cluster-pdp, Panel B].

Analyses for carrying capacity, source-to-sink ratio and total connectivity strength
indicate that their influence on its categorical assignment is concentrated at the low range
of predictor values [@fig-ts-cluster-pdp]. This aligns with the right-skewed distributions
of values across reefs for these three predictor variables. Beyond this lower range, the
partial dependence for these reef characteristics becomes near constant, contrary to their
feature importance scores which suggest these characteristics do contribute important
information to predictions. The contrasting outcomes indicate that interactions between
other reef characteristics are driving behaviour rather than a strong influence from a
single individual characteristic.

>>>>

Long-term reef viability requires not only coral survival and recolonization, but also
maintenance of the physical reef structure through positive carbonate budgets. Scleractinian
corals build the primary habitat framework, but reef persistence depends on the balance
between carbonate accumulation [@perry2015] and loss through bioerosion, physical damage
[@kenyon2023], and dissolution [@eyre2014; @eyre2018]. This balance is influenced by coral
cover and community complexity [@perry2012; @desbiens2021], environmental processes, and
ocean chemistry [@perry2015; @eyre2018]. For GBR mid-shelf and offshore reefs, maintaining
positive carbonate budgets requires approximately 17% live coral cover [@desbiens2021].

@@

![Map of the Great Barrier Reef region, including GBRMPA Management areas (polygons) and all
considered reefs (points). Colour-coded polygons outline GBR marine park management regions,
where orange represents the Far Northern management region, green represents
Cairns/Cooktown, red covers Townsville/Whitsunday and blue covers Mackay/Capricorn
[@GBRMPA2007]. Reef centroids are marked individually for visual clarity and are coloured
according to its bioregion [@GBRMPA2024]. All 2429 reefs included in this investigation are
represented in this figure. Annotations represent major coastal settlements along the
GBR.](../figs/region_map.png){#fig-region-map fig-align="center" width="14.8cm"}

@@

These GCMs were selected based on evaluation conducted by Grose et al., [-@grose2023],
indicating their ability to recreate historical atmospheric temperature and precipitation
conditions in the Australasian region. They span a range of Equilibrium Climate Sensitivity
(ECS) values within the Intergovernmental Panel on Climate Change (IPCC) assessed likely
range. The IPCC defines ECS as the long-term warming caused by a doubling of carbon dioxide
above its pre-industrial concentration [@forster2021]. While this assessment does not
consider oceanic data, Grose et al. [-@grose2023] provide a region-specific selection
mechanism which we use as an indicator of general performance of GCMs for the region. An
overview of the selected GCMs is provided in @tbl-gcms. Note here that Grose et al.,
[-@grose2023] assessed GFDL-ESM4, whereas Dixon et al., [-@dixon2022; -@dixon2023] provides
data based on GFDL-CM4. This substitution is assumed appropriate given they share the same
fundamental representation of the physical climate system.

Shared Socio-economic Pathway (SSP) 2-4.5 was used to represent the plausible emission
pathway. It represents a moderate emissions scenario with continued CO~2~ emissions
comparable to present day until 2050, after which emissions gradually decline
[@hausfather2020; IPCC -@climate2023]. The selected GCMs represent 2.5°C to 4.66°C of global
atmospheric warming in the 2081-2100 period relative to 1850-1900. The use of these GCMs
therefore allow for a range of possible climate futures to be explored, capturing
uncertainty in future responses to CO~2~ increases (@sfig-ecs; @fig-methods-dhws;
[@grose2023; @shepherd2018]) and enables the assessment of modelled coral reef behaviour
over a broad range of independently derived climate futures [@abramowitz2019].

Monthly mean SST climatological baselines, for each 1km^2^ pixel, were calculated from the
daily SST data following NOAA methodology [@liu2017; @skirving2020]. For each pixel of SST
data, the Monthly Mean SST climatology was calculated by computing the mean SST value for
each calendar month (January through December) across the 28-year baseline period from 1985
to 2012. For each month separately, this climatological baseline was adjusted to centre at
year 1988.2857, which attempts to account for any long-term trends within the baseline years
themselves (see Skirving et al. [@skirving2020]). The adjustment ensures that the
climatology represents conditions at a specific temporal reference point rather than being
influenced by gradual warming or cooling trends that may have occurred during the 1985-2012
period. The resulting monthly climatologies serve as a standardized reference against which
current SST conditions can be compared to identify thermal anomalies and assess coral
bleaching risk, following the methodology established by Skirving et al. [-@skirving2020].

@@

| Global Climate Model source | ECS value (°C)     | Interpreted relative severity | Results found in |
|-----------------------|-----------------------|------------------------|-----------------------|
| ACCESS-CM2                  | 4.66 [@grose2023]  | High                          | [@grier2025]     |
| EC-Earth3-Veg               | 4.30 [@wyser2020]  | High                          | [@grier2025]     |
| GFDL-CM4                    | 4.1 [@sentman2025] | High                          | [@grier2025]     |
| ACCESS-ESM1-5               | 3.87 [@ziehn2020]  | Intermediate                  | Main text        |
| NorESM2-MM                  | 2.5 [@seland2020]  | Low                           | [@grier2025]     |

: Information for each GCM used in this investigation including Equilibrium Climate
Sensitivity (ECS) value, interpreted relative severity and where results for an individual
GCM results are found. GCMs in context with assessed plausible ranges of ECS values are
displayed in [@sfig-ecs]. ACCESS-ESM1-5 has an ECS closest to the midpoint of the
very-likely ECS range and was taken as an intermediate model with results displayed in the
main text. {#tbl-gcms}

@@

An exploratory scenario modelling approach was adopted for this investigation. The method
involves exploring a wide range of plausible futures under different assumptions of
environmental (climatic) conditions and ecological factors. Exploratory scenario modelling
embraces uncertainty as an inherent characteristic of complex environmental systems
[@lempert2010] and is a core methodology underpinning the development of robust and adaptive
management strategies [@haasnoot2013].

Exploratory modelling approaches have gained favour in recent years as precise long-term
predictions of future conditions are often impossible in complex systems like coral reefs.
Multiple interacting and compounding sources of uncertainty, and a lack of broad agreement
as to their likelihood, create inherently uncertain outcomes. Rather than attempting to
forecast specific outcomes, exploratory modelling aims to systematically explore many
possible futures to identify patterns. Allowing for the identification of robust management
strategies that perform well across those diverse conditions, as well as conditions in which
such strategies may fail. Such an exploratory approach lends itself well to the purpose of
the investigation. Specifically, to assess the role of thermal stress, connectivity, and
depth to identify the environmental and reef characteristics that support structural
resilience.

@@

To apply exploratory scenario modelling, we employed a modelling framework involving two
separate systems: ADRIA (the Adaptive Dynamic Reef Intervention Algorithms platform) and
CoralBlox. ADRIA is a multi-criteria decision support platform designed to help reef
managers address climate change challenges in coral reef systems [@iwanaga2023]. The ADRIA
platform provides functionality for scenario generation and analysis as well as providing
the technical infrastructure to couple environmental and ecosystem representations.
CoralBlox is a computationally efficient coral population model. It simulates
functional/morphological groups of corals over their life cycles, competing for available
space over time [@ribeirodealmeida2024]. The coupled framework is referred to as
ADRIA-CoralBlox.

The coral demographic representation and associated model implementation are further
detailed in the supplementary material. Of important note, however, is that the
representation of bleaching mortalities followed Bozec et al., [-@bozec2022] albeit with a
scaling coefficient to incorporate the amelioration effect that depth may have, following
Baird et al., [@baird2018]. This depth-scaling reduces the proportion of coral colonies
impacted by high heat stress as depth increases. The impacts of tropical cyclones and wave
activity on reefs were not included in this investigation.

@@

\*\* need information on generation of GBR1 connectivity used \*\*

Connectivity data from 24 days in total were used in this investigation to estimate coral
larval flows. These data span six years from 2015 to 2020. The years 2015, 2016, 2018 and
2019 are each represented by connectivity matrices covering spawning periods over three days
in each year (12 days in total), while 2017 and 2020 are each represented by six spawning
days each, coming to a total of 24 days of connectivity matrices. Indicative connectivity
strengths are implemented in ADRIA-CoralBlox as the mean of all available data for each
location (i.e., mean of all 24 spawning days). While the connectivity strengths do not vary
from year to year within each ADRIA-CoralBlox simulation, they do vary between reefs which
are of varying sizes, providing sufficient heterogeneity to capture spatial patterns in
larval dispersal and settlement dynamics. This spatial variability also enables
investigations into the influence of connectivity strength and its role in driving recovery
processes.

@@

-   $S$ is the sum of outgoing and incoming connectivity, indicating the total connectivity
    strength.

@@

The time series clustering was applied over the time frame between 2025 and 2075. Spatially,
the clustering analyses were conducted for each GBR management area and at the scale of each
reefal bioregion (defined in GBRMPA [-@GBRMPA2024]). Analyses were also performed at the
whole-of-GBR scale but were found to lead to identical conclusions about characteristics
influencing cluster cover compared to management area and bioregion scale analyses and so
are not shown here. To compare time series clustering results across GCMs, management area
scale clustering results are used to preserve coarse scale latitudinal differences between
regions, while maintaining ease of comparison and visualisation. These comparisons include
quantification of differences between low and high cover clusters across management regions
under each GCM using the CID between cluster median time series. To assess how different
reef characteristics influence time series clustering at a finer spatial scale, bioregion
results are used to capture patterns occurring within individual groups of reefs. Results of
all analyses at all clustering scales can be found at Grier and Iwanaga [-@grier2025] for
those interested.

@@

Partial Dependence Analysis was conducted to understand the relationship between individual
features and bioregion-scale time series cluster predictions using a trained Random Forest
(RF) classifier [@friedman2001]. The analysis utilised the reef characteristics of interest
as the predictors, including the median reef depth, mean DHW for the given GCM, total
connectivity strength, and the outgoing-to-incoming connectivity ratio. The full dataset
consisted of the time series cluster assignment of each of the 2429 reefs under the five
GCMs explored, for a total of 12,145 samples. The dataset was randomly split into training
(60%) and testing (40%) subsets to ensure model generalisability. The RF classifier achieved
99% accuracy on the training set and 66% on the test set.

While 99% accuracy suggests the classifier is overfit to the training data, the 66% test
accuracy indicates a meaningful relationship between modelled reef characteristics and their
assigned clusters has been captured. A 3-fold cross-validation was also conducted, arriving
at a similar 67% accuracy rate. The consistent level of accuracy across test and
cross-validation demonstrates the generalised performance of the RF classifier is reliable.
Furthermore, F1 performance scores for Low, Medium, and High clusters were determined to be
0.7, 0.57 and 0.68 respectively, indicating balanced performance across the cluster
categories. Lower performance with the Medium cluster reflects the inherent difficulty of
classifying transitional states in natural systems. For management applications, this level
of performance provides valuable decision support when combined with expert knowledge and
the model's probabilistic outputs.

Partial dependence analysis was performed by applying the trained model to the test dataset.
Values for individual predictors were systematically (one-at-a-time) varied across a grid of
50 equally spaced points spanning the predictor's range, while all other features were left
at their modelled values. The procedure creates (1) a modified test dataset, where the
target predictor was set to the grid value for all samples, (2) probabilistic predictions
for all three cluster classes using the trained RF classifier, and (3) the mean predicted
probability for each class across all test observations. The median was also assessed (not
shown) but was found not to substantially influence the interpretation of results. The
process isolates the marginal effect of each predictor by averaging over the natural
distribution of all other representative reef characteristics in the test dataset.

@@

Using coral cover time series for each reef we calculated the number of years each reef
maintains coral cover above a given percent threshold. The threshold represents the level of
live coral cover necessary to maintain a positive carbonate budget (denoted here as
$\theta_{0}$ to indicate the budget equilibrium). Greater number of years above $\theta_{0}$
suggests greater resilience and opportunity for recovery.

@@

The relationships between the number of years above $\theta_{0}$ and characteristics such as
median reef depth and Log-10 total connectivity strength were then assessed with Spearman's
rank correlation. Spearman's correlation coefficient was adopted as reef depth and
connectivity strength distributions are expected to be non-normal and highly skewed.
Furthermore, reefs are naturally of different types and experience different conditions, and
so the monotonic trends (whether the relationship consistently increase or decrease) rather
than the exact quantitative relationship was of interest for this analysis.

@@

There were substantial differences in the maintenance of coral cover post 2030 between the
GCMs. Figure [-@fig-gcm-cover-trajectories] illustrates the temporal behaviour of reefs
across clusters (Low, Medium and High).

Of the GCMs explored, NorESM2-MM had the lowest Equilibrium Climate Sensitivity (2.5°C) and
reefs exhibit continuing recovery and decline dynamics up to 2075 under this hypothetical
climate future. Unsurprisingly, recovery phases were predominantly seen in the reefs in the
High category, with Medium and Low exhibiting lesser probability of such periods of
recovery. Generally, recovery and increases in coral cover were visible in simulated results
up to 2045. Under NorESM2-MM, the Far Northern Management area were found to have the most
similar temporal behaviour between Low and High categories compared to other Management
areas (@tbl-cluster-cid) due to the low degree of coral cover overall
(@fig-gcm-cover-trajectories).

Intermediate levels of coral cover maintenance can be seen in ACCESS-ESM1-5 (ECS value of
3.87°C) when compared to the other GCMs (of higher and lower ECS values). Under these
intermediate conditions (in relative terms), there is potential for reefs in the High
category to maintain coral cover levels around 20% to 2050. The highest peak DHW levels
occur under GFDL-CM4 between 2045 and 2050, while greater increases in overall DHWs time
series occur under ACCESS-CM2 (@fig-methods-dhws). Under conditions represented by high ECS
value GCMs - specifically, ACCESS-CM2 (4.66°C), EC-Earth3-Veg (4.30°C) and GFDL-CM4
(4.1°C) - all reefs are largely in decline across all management areas from 2055
(@tbl-cluster-cid).

@@

Comparing across GCMs, it is common for reefs to switch between categories, including from
Low to High and vice versa (@sfig-gcm-bioregion-cluster-assignment). Of 2429 reefs, 20%
(503) of reefs have consistent category assignment across all GCMs. Of the 503 consistently
assigned reefs, 49% are within the Far Northern Management area, with 7%, 13% and 31%
occurring in the Cairns-Cooktown, Townsville-Whitsunday and Mackay-Capricorn Management
areas respectively (@sfig-gcm-cluster-map). Of the 503 consistently assigned reefs, 32%, 7%
and 61% are clustered into Low, Medium and High clusters respectively. High levels of
assignment variability across GCMs are expected as due to spatial differences in each given
GCM’s DHW trajectories.

```{julia}
#| echo: false
#| warning: false
#| message: false
#| results: asis
#| label: tbl-cluster-cid
#| tbl-cap: Complexity invariant distance (CID) between low and high cluster median time series in each management area under each GCM. CID values are calculated for percentage cover time series (@fig-gcm-cover-trajectories) and Degree Heating Week (DHW) time series (@sfig-gcm-dhw-trajectories). Complexity invariant distance represents the Euclidean distance between time series, corrected by the complexity of the two time series. CID values are used to quantify overall differences between these two cluster levels at management area and GCM levels. CID values for percentage cover and DHW time series are not directly comparable as the time series use different units.

@@

In contrast to the broad temporal behaviour at the Management area scale, the relationship
between reef characteristics and assigned categories were found to be similar at the
bioregion scale across GCMs. Results are only shown and discussed for ACCESS-ESM1-5, an
intermediate ECS scenario (@sfig-ecs). Those interested in results for the other assessed
GCMs are referred to Grier and Iwanaga [-@grier2025]. Across bioregions, reefs in the High
category not only had greater levels of coral recovery but displayed persistence to 2050
(@sfig-access-esm1-5-bioregion-timeseries). Reefs in the High category were consistently
found to have the greatest median reef depths across bioregions
(@fig-access-esm1-5-bioregion-depth).

@@

In many bioregions reefs in the Low category have the lowest correlation between coral cover
and inverse-DHW time series whereas reefs in the High category were found to have the
strongest correlation (@sfig-access-esm1-5-bioregion-dhwcor). Across six of the 29
bioregions reef mean DHW levels increased as cluster cover decreased
(@sfig-access-esm1-5-bioregion-dhwmean). There were no consistent patterns across bioregions
in their reefs' initial proportion of habitable area occupied by corals and cluster
assignment (@sfig-access-esm1-5-bioregion-initialproportion). Across most bioregions there
were not consistent differences between clusters in connectivity metrics such as Log-10
source-to-sink connectivity ratio (@sfig-access-esm1-5-bioregion-logsosi) and Log-10 total
connectivity strength (@sfig-access-esm1-5-bioregion-totalconnectivity).

@@

A Partial Dependence Analysis was conducted using a Random Forest classifier to identify
reef characteristics that influence its temporal behaviour, and thus its cluster assignment
into Low, Medium, or High. Bioregion classification was excluded from this analysis as a
predictor due to its categorical nature providing limited interpretable information, and is
further justified by its low feature importance score of 10%, the lowest of all predictors
(@fig-ts-cluster-pdp; Panel A). Reef median depth and mean DHW level have the highest
feature importance scores, accounting for 25% and 23% predictive importance respectively.
The carrying capacity of a reef accounts for 14.3% feature importance. Connectivity strength
and source-to-sink ratio both had feature importance scores of \~13%.

Reefs have the greatest probability of being clustered into Low if shallow, and High at
greater depths, with medium cluster reefs exhibiting a lower, and more stable, assignment
probability across depth ranges (@fig-ts-cluster-pdp; Panel B). A large increase in the
probability of a reef being categorised as High can be seen between 10 and 15 meters depth.
This is at the edge of the depth range used in current deployment efforts [@howlett2021;
@randall2021; @whitman2024] but indicates a depth range that could be adopted to maximise
longevity and benefits of, for example, coral seeding activities. Across mean DHW values,
assignment probability is roughly equal across the three categories up to 6 DHW. A sharp
delineation can be seen \> 6 DHW when comparing High with other categories, with lower
assignment probability (\~25%) compared to Medium and Low (\~40%) (@fig-ts-cluster-pdp;
Panel B).

Analyses for carrying capacity, source-to-sink ratio and total connectivity strength
indicate that their influence on its categorical assignment is concentrated at the low range
of predictor values (@fig-ts-cluster-pdp). This aligns with the right-skewed distributions
of values across reefs for these three predictor variables. Beyond this lower range, the
partial dependence for these reef characteristics becomes near constant, contrary to their
feature importance scores which suggest these characteristics do contribute important
information to predictions. The contrasting outcomes indicate that interactions between
other reef characteristics are driving behaviour rather than a strong influence from a
single individual characteristic.
